---
title: "Covid Project"
date: "2025-03-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Overview
Goal is to do a short data exploration of world wide Covid data.  The analysis and modeling will focus on total Covid cases and deaths. 


```{r GITHUB_URLS, message = FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names = c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_confirmed_US.csv")

urls = str_c(url_in, file_names)
urls
```

```{r import_data, message = FALSE}
global_cases = read_csv(urls[1])
global_deaths = read_csv(urls[2])
us_cases = read_csv(urls[3])
us_deaths = read_csv(urls[4])
```

```{r global_tidy_data, message = FALSE}
global_cases <- global_cases %>%
  pivot_longer(cols= -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols= -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

```

```{r global}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

summary(global)

global <- global %>% filter(cases > 0)

summary(global)
```

```{r us_tidy_data, message = FALSE}
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

US <-us_cases %>%
  full_join(us_deaths)

```

```{r global2}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ",",
        na.rm = TRUE,
        remove = FALSE)

```

```{r lookup}
uid_lookup_file = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid = read_csv(uid_lookup_file)%>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```



```{r Global_by_country}
Global_by_cntry <- global %>%
  group_by( Country_Region, date) %>%
  # add up counties and population
  summarize(cases = sum(cases), 
            deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select( Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```

```{r graph1}
Global_total <- Global_by_cntry %>%
  group_by(date) %>%
  summarize(cases = sum(cases))

ggplot(Global_total, aes(x = date, y = cases)) +
  geom_line(color = "blue") +
  labs(
    title = "Global Total Number of Cases Over Time",
    x = "Date",
    y = "Total Cases"
  )
```

```{r graph2}
Global_total_deaths <- Global_by_cntry %>%
  group_by(date) %>%
  summarize(deaths = sum(deaths))

ggplot(Global_total_deaths, aes(x = date, y = deaths)) +
  geom_line(color = "red") +
  labs(
    title = "Global Total Number of Deaths Over Time",
    x = "Date",
    y = "Total Deaths"
  ) 
```

```{r model_data_prep}
global_totals <- Global_by_cntry %>%
     group_by(date) %>%
     summarize(
         total_cases = sum(cases, na.rm = TRUE),
         total_deaths = sum(deaths, na.rm = TRUE),
         total_population = sum(Population, na.rm = TRUE),
         .groups = "drop"
     )

global_totals  %>%
     ggplot()+
     geom_point(aes(x=date, y=total_deaths), color = "red") +
     geom_point(aes(x=date, y=total_cases), color = "blue")+
     geom_point(aes(x=date, y=total_population), color = "purple")

```


## Univariate Model using total cases
Build a quick linear model using only total cases to predict total deaths.
```{r model_tot_deaths}
mod <-lm(data=global_totals, total_deaths ~ total_cases )
summary(mod)

global_totals <- global_totals %>%
  mutate(predicted_deaths = predict(mod, newdata = global_totals))

global_totals  %>%
     ggplot()+
     geom_point(aes(x=date, y=total_deaths), color = "blue") +
     geom_point(aes(x=date, y=predicted_deaths), color = "red") +
    labs(
    title = "Modelled Deaths vs Actual Deaths Over Time",
    x = "Date",
    y = "Total Deaths"
  ) 
```

## Model starting with July 2020
This enables adding population as in input to the model, since prior the population metric didn't look correct in early 2020.
```{r model_tot_deaths2}
 global_totals <- global_totals %>%
     filter(date >= as.Date("2020-07-01"))

mod <-lm(data=global_totals, total_deaths ~ total_cases + total_population )
summary(mod)

global_totals <- global_totals %>%
  mutate(predicted_deaths = predict(mod, newdata = global_totals))

global_totals  %>%
     ggplot()+
     geom_point(aes(x=date, y=total_deaths), color = "blue") +
     geom_point(aes(x=date, y=predicted_deaths), color = "red")
```

## Conclusion
Based on the models, population and Covid cases are good predictors for Covid deaths. The second model tried to work around the inaccurate early 2020 population data, but is still impacted by a small population difference in mid-2022.  With more time, it would be interesting to do a similar analysis by continents.  Decided not to, since the country ISO look up data didn't contain a continent variable. 
