---
title: "NYPD_project"
date: "2025-03-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#library(lubridate)
#library(ggplot2)
```

The goal of this project is to explore the NYPD shooting data by creating a few visuals and a model.

Import Data:
```{r get_data}
url_nypd <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_data <- read.csv(url_nypd)
#summary(nypd_data)
```

Clean the data by limiting to variables of interest, converting occur date and time to actual date and time variables.
```{r clean_data}
nypd_clean <- nypd_data %>%
  select(c("OCCUR_DATE","OCCUR_TIME", "BORO","PRECINCT",
           "STATISTICAL_MURDER_FLAG","VIC_AGE_GROUP","VIC_SEX","VIC_RACE"
           , "PERP_AGE_GROUP", "PERP_SEX", "PERP_RACE" )) %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_TIME = hms(OCCUR_TIME),
         STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG),
         Shootings = 1,
         Year = year(OCCUR_DATE),
         Month = month(OCCUR_DATE),
         Hour = hour(OCCUR_TIME))
```

Various graphs looking at the number of shootings by year, hour, boroughs, precincts, perp data, and killings.  
```{r graph1}
nypd_clean %>%
  ggplot(aes(x = Year)) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Year",
       x = "Year",
       y = "Total Number of Shootings")
```

```{r graph_hr}
nypd_clean %>%
  ggplot(aes(x = Hour)) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Hour",
       x = "Hour",
       y = "Total Number of Shootings")
```

```{r graph2}
nypd_clean %>%
  ggplot(aes(x = BORO, fill = BORO)) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Borough",
       subtitle = "Years: 2006 - 2021",
       x = "Boroughs",
       y = "Total Number of Shootings")

```

```{r graph3}
nypd_clean %>%
  ggplot(aes(x = PRECINCT)) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Precinct",
       x = "Precinct",
       y = "Total Number of Shootings")
```


```{r graph4}
nypd_clean %>%
  ggplot(aes(x = PERP_RACE)) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Victims Race",
       x = "Victims Race",
       y = "Total Number of Shootings")
```

```{r graph5}
nypd_clean %>%
  ggplot(aes(x = PERP_AGE_GROUP )) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Victims Age Group",
       x = "Victims Age Group",
       y = "Total Number of Shootings")
```

```{r graph6}
nypd_clean %>%
  ggplot(aes(x = PERP_SEX )) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Victims Sex",
       x = "Victims Sex",
       y = "Total Number of Shootings")
```

```{r graph7}
nypd_clean %>%
  ggplot(aes(x = STATISTICAL_MURDER_FLAG  )) +
  geom_bar() +
  labs(title = "NYPD Shooting Incidents by Murders",
       x = "Murders",
       y = "Total Number of Shootings")
```

As we saw with hour there is a quadradic relationship.  So, let's try to a simple model of shooting by  hour.
```{r model}
nypd_hour_mod <- nypd_clean %>%
     group_by(Hour, Shootings) %>%
     summarize(Shootings = sum(Shootings),
               STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG))

mod <-lm(data=nypd_hour_mod, Shootings ~ Hour )
summary(mod)

nypd_hour_mod <- nypd_hour_mod %>% 
     ungroup() %>% 
     mutate(pred = predict(mod, newdata = nypd_hour_mod))
 nypd_hour_mod <- nypd_hour_mod %>% 
     rowwise() %>% 
     mutate(pred = predict(mod, newdata = cur_data()))
```

```{r model_graph}
nypd_hour_mod  %>%
  ggplot()+
  geom_point(aes(x=Hour, y=Shootings), color = "blue") +
  geom_point(aes(x=Hour, y=pred), color = "red")
```

As we saw with model one, a linear relationship doesn't model shootings well.  Let's try adding a quadratic hour variable and see if that gets a better model.  

```{r model_2}
nypd_hour_mod <- nypd_clean %>%
     group_by(Hour, Shootings) %>%
     summarize(Shootings = sum(Shootings),
               STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG)) %>% 
     mutate(hour_sq = Hour^2)

mod <-lm(data=nypd_hour_mod, Shootings ~ Hour + hour_sq )
summary(mod)

nypd_hour_mod <- nypd_hour_mod %>% 
     ungroup() %>% 
     mutate(pred = predict(mod, newdata = nypd_hour_mod))
 nypd_hour_mod <- nypd_hour_mod %>% 
     rowwise() %>% 
     mutate(pred = predict(mod, newdata = cur_data()))
```


```{r model_graph2}
nypd_hour_mod  %>%
  ggplot()+
  geom_point(aes(x=Hour, y=Shootings), color = "blue") +
  geom_point(aes(x=Hour, y=pred), color = "red")
```

Conclusion

Looking at the charts, we see shootings were going down until 2019 and then there is a spike and potential re-normalization in 2020 with COVID.  There is a clear relationship between time (hour) and shootings.  From the other charts there are some other metrics that could have a relationship with a shoot, such as which borough and sex of perp.

My model is a simple linear model using time (hour) to predict number of shootings.  The relationship isn't linear, so hour is squared.  Then a really strong model is produced. 